Model Order Selection and Eigen Similarity based Framework for Detection and Identification of Network Attacks

### 2016-11-02 docs/papers/2016JNCA/2016JNCA.tex
several updates
baseline para as modificacoes do paper JNCA
updates of introduction, related work and dataset description

\begin{abstract}
Novel schemes for attack detection are crucial to identify adaptive malicious traffic coming from sources that are quickly mobilized by attackers in high throughput communication networks. In this context, signal processing techniques have been applied to attack detection due to their capability to detect anomalies that are previously unknown, i.e. blind detection. This paper proposes a signal processing framework for the detection and identification of network attacks using concepts of model order selection (MOS), eigenvalues and similarity analysis. In order to validate the proposed framework, we consider a network traffic sample that contains malicious activity such as denial of service (DoS) and port probing. We propose to model the network traffic as a superposition of components, namely, user's operations (legitimate traffic), network service operation not related to the user (noise) and the malicious activity. The experiments performed in a real network show that the proposed blind detection approach achieves satisfactory levels of accuracy in terms of timely detection and identification of TCP/UDP ports under attack.
\end{abstract}

@@

Information security may consist of both technical and procedural aspects. The former includes equipment and security systems, while the latter corresponds to security rules and recommendations. Intrusion detection and intrusion prevention systems are security systems used, respectively, to detect (passively) and prevent (proactively) threats to computer systems and computer networks. Such systems can work in the following fashions: signature-based, anomaly-based or hybrid \cite{Huang2009,mudzingwa2012study}.

In the context of anomaly-based schemes, this work proposes an approach for the detection of malicious traffic in computer networks. Inspired by \cite{david2011blind,da2012improved}, this work models the network traffic using a signal processing formulation as a composition of three components: legitimate traffic, malicious traffic and noise, taking into account the incoming and outgoing traffic in certain types of network ports (TCP or UDP). To the best of our knowledge there is no similar model in the literature. The proposed technique is based on eigenvalue analysis, model order selection (MOS) and similarity analysis. In contrast to \cite{david2011blind,da2012improved,tenorio2013greatest}, MOS and eigenvalue analysis are applied to detect time frames under attack. In addition, we also evaluate the accuracy of the proposed framework. Additionally, this proposed approach has its accuracy evaluation based on eigen similarity analysis for extracting detailed information about accurate time and network ports under attack.

@@

Several methods have been proposed for the identification and characterization of malicious activity in computer networks. Classical methods typically employ data mining \cite{he2008applying,ghourabi2010data} and regular file analysis \cite{raynal2004honeypot} to detect patterns that indicate the presence of specific attacks in network traffic.

Data mining is often used to describe the process of extracting useful information from large databases. Multiple methods of data mining are used in \cite{he2008applying} to analyze data flow in a network with the aim of identifying characteristics of malicious traffic in large scale environments. Researchers have applied data mining techniques in log analysis \cite{ghourabi2010data} to improve intrusion detection performance. However, data mining techniques used so far in network analysis require prior collection of large data sets, which is a limitation of several schemes for online analysis \cite{Huang2009}.

@@

Principal Component Analysis (PCA) is a statistical technique commonly used for dimensionality reduction. It uses an orthogonal transformation to convert a set of correlated variables into a set of linearly uncorrelated variables, where the first principal components have the largest variance. PCA has been used in attack detection \cite{almotairi2009technique}. However, PCA requires human intervention in order to identify abnormalities based on the eigenvalues profiles, if used without complementary techniques. Besides being prone to higher errors and false positives, such human intervention makes PCA not useful for real time applications. Therefore, in order to automate the analysis of eigenvalues profile, model order selection (MOS) schemes should be incorporated.

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work did not address problems without significant outliers, such as port scan attacks. Zonglin \emph{et al} \cite{Zonglin2009} proposed a method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al.} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work was not applied to probing and denial of service (DoS) attack detection, simultaneously.

@@

Lee \emph{et al.} \cite{Lee2013} proposed OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. In contrast to Lee \emph{et al.}, the framework applies MOS for detection of time frames under attack and similarity analysis to extract details for detection of time and ports under attack. Additionally, Lee \emph{et al.} only evaluated their proposed scheme for covariance analysis, while we adopted an analysis based on sample covariance of zero mean variables and sample covariance of zero mean and unitary standard deviation variables, for DoS and probing attacks, respectively.

The proposed framework does not require either a significant amount of logs to detect attacks, nor prior data collection, in order to make comparisons and evaluate the existence of malicious traffic. The proposed solution is automatic and blind for detection of time frames under probing and DoS attacks through MOS and eigen analysis. Moreover, we apply eigen similarity analysis to identify details of time and ports under network attacks.

@@

This section presents details of the simulated scenario along with a description of the dataset model as a signal superposition of legitimate traffic, noise and malicious traffic. Subsection \ref{sec:DataCollection} describes the environment and scenario adopted in order to reproduce DoS and probing attacks. Subsection \ref{sec:ModelingData} presents how network traffic can be modeled as signal superposition, and Subsection \ref{sec:AnalyzedScenario} details the traffic of synflood, fraggle and port scan attacks.

@@

The environment of the analyzed scenario is composed of two computers and one router with access to the Internet and to an internal network, where the simulation of legitimate traffic, noise, DoS and port scan attacks are performed. During the traffic generation, one computer assumes the role of the attacker, while the other is the victim, according to scenario represented by Figure \ref{fig:fig1}.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=5cm, width=7cm]{results/figures/fig09.png}
     \caption{Scenario to reproduce legitimate traffic, noise, DoS and port scan.}
     \label{fig:fig1}
\end{figure}

@@

By modeling the dataset as a signal superposition, the network traffic (\textbf{X}) can be characterized as a mixture of three components: legitimate traffic (\textbf{S}), noise (\textbf{N}) and malicious traffic (\textbf{A}), according to the following expression:

@@

The network attacks evaluated by this work are: synflood, fraggle and port scan. The first two attacks can be qualified as DoS attacks, while the last one can be qualified  as probing or port scanning attack.

@@

In TCP SYN scan, a SYN packet is sent to the destination and two types of responses may occur: SYN/ACK or RST/ACK. In the first case, the destination port is in the listening state, in the second case, the destination port is not listening. At the end of each port scanning, a RST/ACK packet is sent by the system that is performing the port  scan. Therefore, a full connection or a complete three-way handshake is never established, which makes the detection of the attack sender more difficult, and requires approaches able to identify probing attacks without connection stablishment. The UDP scan technique sends UDP packets to the destination port, and if it responds with a \emph{ICMP port unreachable} message, then it indicates that the scanned port is closed. On the other hand, if a message is not received, then the port is considered as open.

@@

@@

\begin{figure}[h!]
	\centering
     \includegraphics[height=11cm, width=9cm]{mos_eigen_similarity.eps}
     \caption{Overview of The Framework for Detection and Identification of Network Attacks.}
     \label{fig:fig80}
\end{figure}

@@

According to DoS and port scan attacks' behavior, DoS attacks and port scan attacks can be characterized as covariance aware attack \citep{jin2004covariance} and correlation aware attack \citep{lakhina2005mining}, respectively. These characteristics are substantiated by the results obtained through the analysis based on sample covariation of zero mean variables and on covariance of zero mean and unitary standard deviation variables, described in Section \ref{sec:experimentalresults}, which shows that the main components of DoS attacks are dominated by the variables with more variance and that the traffic associated with port scan attack does not generate many logs, however, it presents high covariance of zero mean and unitary standard deviation variables.

Therefore, to detect DoS attacks, it is necessary to calculate the sample covariance matrix $\boldsymbol{\hat{R}}_{yy}^{(q)}$ of the zero mean samples given by

@@

Once the $\boldsymbol{\hat{R}}_{yy}^{(q)}$ and $\boldsymbol{\hat{R}}_{zz}^{(q)}$ have been obtained for DoS and port scan attack detection, respectively, and since the next steps are the same for both sample covariance matrices, we refer to $\boldsymbol{\hat{R}}_{yy}$ and $\boldsymbol{\hat{R}}_{zz}$ as a matrix $\boldsymbol{C}$. Therefore, the following step of the algorithm is the eigenvalue decomposition (EVD), calculated according to (\ref{eq:eq06}), in order to obtain the vector of eigenvalues $\boldsymbol{e}^{(q)}$ associated with each matrix, according to (\ref{eq:eq060}).

\begin{equation}\label{eq:eq06}
\boldsymbol{C}^{(q)} = \boldsymbol{V}^{(q)}\boldsymbol{\Lambda}^{(q)}\boldsymbol{V}^{(q)^{\rm T}},
\end{equation}
\begin{equation}\label{eq:eq060}
\boldsymbol{e}^{(q)} = \rm diag(\boldsymbol{\Lambda}^{(q)}),
\end{equation}

@@

\begin{equation}\label{eq:eq11}
s_n = \frac{\abs{\boldsymbol{v}^{(q)} \cdot \boldsymbol{v}_{(n)}}}{\norm{\boldsymbol{v}^{(q)}}\norm{\boldsymbol{v}_{(n)}}},
\end{equation}
where $s_n$ denotes the absolute similarity degree of the $n$-th minute, $\boldsymbol{v}^{(q)}$ is the most significant eigenvectors of a selected set of minutes without network attack, and $\boldsymbol{v}_{(n)}$ is the most significant eigenvectors obtained after append the target $n$-th minute of traffic to be performed the DoS and port scan attack identification.

@@

This section presents the performed experiments and the acquired results. First, in Section \ref{sec:AnalyzedScenario}, the scenario adopted in the experiments is summarized. Then, Section \ref{sec:largesteigenvaluesanalysis} shows the results of the largest eigenvalue analysis by time frames. In Section \ref{sec:MOSSchemesEvaluation} are described the results of the evaluated MOS schemes for attack detection. Section \ref{sec:EigenvalueAnalysis} presents the results of the eigenvalue analysis for identification of time frames under attack, Section \ref{sec:EigenSimilarityAnalysis} shows the results of similarity analysis for detailed DoS and port scan identification, and Section \ref{sec:Complexity} discuss the computational complexity of the proposed framework.

@@

For the evaluation of MOS Schemes acuracy for DoS and port scan detection, the framework defines that it is necessary to obtain the largest eigenvalue of each time frame, through eigen decomposition from a covariance of zero mean variables or covariance matrix of zero mean and unitary standard deviation variables, calculated from the evaluated traffic, as described in Section \ref{sec:prop_getv}. Through eigenvalue analysis of traffic with DoS or port scan attacks, it is possible to visualize a significant difference between the largest eigenvalues and the remain eigenvalues, which can indicate a relationship between an outlier and time frames under attack.

@@

These results highlight that all $q$-th time frames where a network attack was simulated, present high significant variance between the largest eigenvalue and the remaining eigenvalues, obtained from sample covariance matrix, for DoS detection, or from covariance matrix of zero mean and unitary standard deviation variables, for port scan detection. Therefore, we propose to apply the vector of the largest eigenvalues to MOS schemes in order to evaluate their accuracy for identification of time frames under attack, motivated by the fact that it is relevant to apply MOS schemes to automate the attack detection process, taking into account the characteristics of the evaluated eigenvalues.

@@

\begin{table}[h!]
  \centering
  \scriptsize
  \caption{MOS schemes applied to port scan and DoS detection}
  \label{tab:tab4}
  \begin{tabular}{ c c c c c c c c }
	\toprule
	\multirow{2}{*}{\textbf{Type of analysis} $q$} &\multicolumn{6}{c}{\textbf{MOS schemes (estimated model order $\hat{d}$)}} &{\textbf{(d)}}\\ 
			\hhline{~------~}
		&\textbf{AIC} &\textbf{MDL} &\textbf{EDC} &\textbf{RADOI} &\textbf{EFT} &\textbf{SURE}\\
	\midrule
	Detection of synflood \\(presence of attack) &2 &1 &\textbf{1} &5 &\textbf{1} &4 &\textbf{1} \\
	Detection of synflood \\(absence of attack) &1 &1 &\textbf{0} &1 &\textbf{0} &3 &\textbf{0} \\
	\midrule
	Detection of fraggle \\(presence of attack) &1 &1 &\textbf{1} &5 &\textbf{1} &4 &\textbf{1} \\
	Detection of fraggle \\(absence of attack) &1 &1 &\textbf{0} &1 &\textbf{0} &3 &\textbf{0} \\
	\midrule
	Detection of port scan \\(presence of attack) &1 &1 &\textbf{1} &1 &\textbf{1} &9 &\textbf{1} \\
	Detection of port scan \\(absence of attack) &0 &0 &\textbf{0} &1 &\textbf{0} &1 &\textbf{0} \\
	\midrule
	Detection of synflood/fraggle \\(presence of attack) &2 &2 &\textbf{2} &5 &\textbf{2} &5 &\textbf{2} \\
	Detection of synflood/fraggle \\(absence of attack) &1 &1 &\textbf{0} &1 &\textbf{0} &3 &\textbf{0} \\
    \bottomrule
  \end{tabular}
\end{table}

@@

In Table \ref{tab:tab4}, two MOS schemes outperforms from the others, EDC and EFT. Efficient Detection Criterion (EDC) and Exponential Fitting Test (EFT) are the most effective schemes, correctly estimating the number of attacks in comparison to the expected values for effetctive attack detection, as defined by the column of real values in Table \ref{tab:tab4}. The AIC and MDL schemes are satisfactory only for port scan detection, however SURE and RADOI schemes did not show effective results for port scan or DoS detection.

Although EDC and EFT presented the same accuracy on the evaluation, the EDC scheme requires less processing time than EFT, which is an important criteria to select EDC as the MOS scheme for DoS and port scan detection on the remain experiments.

@@

These results highlight the capabilitiy of change detection based on similarity between legitimate and malicious traffic from DoS or port scan attacks, endorsing the effectiveness and safety for adoption of threshold for attack detection through eigen similarity analysis.

@@

Given $\hat{N}$, which is the set of estimated $n$-th minutes under attack, it is possible to apply cosine similarity analysis to identify variation of the most significant eigenvectors, caused by the insertion of anomalous network traffic by a selected $m$-th port, during a $n$-th minute. Therefore, the incremental individualized and individual approaches of eigen similarity analysis were evaluated, for detection of ports under DoS and port scan attacks, according to results presented in following tables. For this evaluation, the $v$ last most significant eigenvectors without attack was used as reference for similarity analysis against each target port $m$-th.

@@

According to the shown significant lack of similarity between legitimate and malicious traffic, it is possible to adopt safe thresholds for DoS and port scan detection through eigen similarity analysis.

@@

This paper evaluated the effectiveness of MOS schemes for fraggle attack detection, extending our previous work \cite{tenorio2013greatest} and showing that the analysis of the largest eigenvalues by time frames can be applied to detect the number of port scanning, and DoS attacks, but still requiring more information for detailed attack detection. Therefore, we proposed a novel approach for detailed network attack detection, based on eigen similarity analysis.

@@

According to the significant similarity difference between legitimate and malicious traffic, it is possible to adopt safe thresholds for DoS and port scan detection through eigen similarity analysis.

>>>>

\begin{abstract}
Novel schemes for attack detection are crucial to identify adaptive malicious traffic coming from sources that are quickly mobilized by attackers in high throughput communication networks. In this context, signal processing techniques have been applied to attack detection due to their capability to detect anomalies that are previously unknown, i.e. blind detection. This paper proposes a signal processing framework for the detection and identification of network attacks using concepts of model order selection (MOS), eigenvalues and similarity analysis. In order to validate the proposed framework, we consider network traffic datasets that contains malicious activity such as flood and port probing attacks. We propose to model the network traffic as a superposition of components, namely, user's operations (legitimate traffic), network service operation not related to the user (noise) and the malicious activity. The experiments performed in a real network and in a well known public dataset show that the proposed blind detection approach achieves satisfactory levels of accuracy in terms of timely detection and identification of TCP/UDP ports under attack.
\end{abstract}

@@

Information security may consist of both technical and procedural aspects. The former includes equipment and security systems, while the latter corresponds to security rules and recommendations. Intrusion detection and intrusion prevention systems are security systems used, respectively, to detect (passively) and prevent (proactively) threats to computer systems and computer networks. Such systems can work in the following fashions: signature-based, anomaly-based or hybrid \cite{Huang2009,mudzingwa2012study}. Additionally, anomaly detection techniques can be categorized in classification, statistical, Information theory and clustering based \cite{bhuyan2014network,ahmed2016survey,osanaiye2016distributed}.

In the context of anomaly-based schemes, this work proposes a statistical approach based on signal processing techniques for the detection of malicious traffic in computer networks. Inspired by \cite{david2011blind,da2012improved}, this work models the network traffic using a signal processing formulation as a composition of three components: legitimate traffic, malicious traffic and noise, taking into account the incoming and outgoing traffic in certain types of network ports (TCP or UDP). To the best of our knowledge there is no similar model in the literature. The proposed technique is based on eigenvalue analysis, model order selection (MOS) and similarity analysis. In contrast to \cite{david2011blind,da2012improved,tenorio2013greatest}, MOS and eigenvalue analysis are applied to detect time frames under attack. In addition, we also evaluate the accuracy and performance of the proposed framework applied to a simulated scenario and to the DARPA 1998 Dataset \citep{osanaiye2016distributed}, which is a well known network traffic dataset. Furthermore, this proposed approach has its accuracy evaluation based on eigen similarity analysis for extracting detailed information about accurate time and network ports under attack.

@@

Several methods have been proposed for the identification and characterization of malicious activity in computer networks. Classical methods typically employ data mining \cite{he2008applying,ghourabi2010data,osanaiye2016distributed} and regular file analysis \cite{raynal2004honeypot} to detect patterns that indicate the presence of specific attacks in network traffic.

Data mining is often used to describe the process of extracting useful information from large databases. Multiple methods of data mining are used in \cite{he2008applying,osanaiye2016distributed} to analyze data flow in a network with the aim of identifying characteristics of malicious traffic in large scale environments. Researchers have applied data mining techniques in log analysis \cite{ghourabi2010data} to improve intrusion detection performance. However, data mining techniques used so far in network analysis require prior collection of large data sets, which is a limitation of several schemes for online analysis \cite{Huang2009}.

@@

Principal Component Analysis (PCA) is a statistical technique commonly used for dimensionality reduction. It uses an orthogonal transformation to convert a set of correlated variables into a set of linearly uncorrelated variables, where the first principal components have the largest variance. PCA has been used in attack detection \cite{almotairi2009technique}. However, PCA requires human intervention in order to identify abnormalities based on the eigenvalues profiles, if used without complementary techniques.

Callegari \emph{et al} \cite{Zonglin2009} proposed a PCA-based method for identifying the traffic flows responsible for an anomaly detected at the aggregate level and evaluated their proposal through a dataset with synthetic anomalies added in the data. However, Callegari \emph{et al} focused in flood attack detection, not addressing probe attack detection, and their approach relies on visual analysis.

Lee \emph{et al.} \cite{Lee2013} proposed OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. In contrast to Lee \emph{et al.}, the framework applies MOS for detection of time frames under attack and similarity analysis to extract details for detection of time and ports under attack. Additionally, Lee \emph{et al.} only evaluated their proposed scheme for covariance analysis, while we adopted an analysis based on sample covariance of zero mean variables and sample covariance of zero mean and unitary standard deviation variables, for flood and probe attacks, respectively.

Besides being prone to higher errors and false positives, such human intervention makes PCA not useful for real time applications. Therefore, in order to automate the analysis of eigenvalues profile, model order selection (MOS) schemes should be incorporated.

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work did not address problems without significant outliers, such as port scan attacks. Zonglin \emph{et al} \cite{Zonglin2009} proposed a method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al.} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work was not applied to probe and flood attack detection, simultaneously.

@@

The proposed framework does not require either a significant amount of logs to detect attacks, nor prior data collection, in order to make comparisons and evaluate the existence of malicious traffic. The proposed solution is automatic and blind for detection of time frames under probe and flood attacks through MOS and eigen analysis. Moreover, we apply eigen similarity analysis to identify details of time and ports under network attacks.

Several approaches for network attack detections uses the KDD 99 \cite{ji2016multi,ahmed2016survey,osanaiye2016distributed,bhuyan2014network} datasets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 dataset are criticized by for the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled datasets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} dataset is the refined version of the KDD 99 dataset that redundant data records are removed, in order to avoid biased classifications. Additionally, some approaches uses simulated \cite{callegari2011novel} scenarios or non-public datasets their evaluations. Since the proposed approach relies on a packet level analysis and the KDD 99 and NSL-KDD datasets adopt a traffic aggregation by connections, we consider the use of a simulated scenario on a real network and the DARPA 1998 dataset, which is the source for the creation of the KDD 99 and NSL-KDD datasets. It is important to note that the proposed approach is not based on learning or classification techniques, which are more susceptible to biased results caused by the issues in the DARPA/KDD datasets.

@@

This section presents details of the simulated scenario and the selected cases of the DARPA dataset, along with a description of the dataset model as a signal superposition of legitimate traffic, noise and malicious traffic. Subsection \ref{sec:DataCollection} describes the environment and scenario adopted in order to reproduce flood and probe attacks. Subsection \ref{sec:ModelingData} presents how network traffic can be modeled as signal superposition. Subsection \ref{sec:SynfloodFraggleandPortscan} details the traffic of synflood, fraggle and port scan attacks, and Subsection \ref{sec:DarpaDataset} discuss the use of the DARPA dataset for evaluation of the proposed approach.

@@

The environment of the analyzed scenario is composed of two computers and one router with access to the Internet and to an internal network, where the simulation of legitimate traffic, noise, flood and port scan attacks are performed. During the traffic generation, one computer assumes the role of the attacker, while the other is the victim, according to scenario represented by Figure \ref{fig:fig1}.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=5cm, width=7cm]{results/figures/fig09.png}
     \caption{Scenario to reproduce legitimate traffic, noise, flood and port scan.}
     \label{fig:fig1}
\end{figure}

@@

By modeling the dataset as a signal superposition, the network traffic (\textbf{X}) can be characterized as a mixture of three components: legitimate traffic (\textbf{U}), noise (\textbf{N}) and malicious traffic (\textbf{A}), according to the following expression:

@@

The network attacks evaluated by this work are: synflood, fraggle and port scan. The first two attacks can be qualified as flood or denial of service (DoS) attacks, while the last one can be qualified as probe or port scanning attack. DoS attempts to block access to system or network resources, it is implemented by either forcing targets to be unavailable through the exploiting of system vulnerabilities, or consuming resources through large amount of network traffic, characterizing flood attacks. Probe attacks scan the networks to to collect information about host, such as IP addresses, ports and services.

@@

In TCP SYN scan, a SYN packet is sent to the destination and two types of responses may occur: SYN/ACK or RST/ACK. In the first case, the destination port is in the listening state, in the second case, the destination port is not listening. At the end of each port scanning, a RST/ACK packet is sent by the system that is performing the port scan. Therefore, a full connection or a complete three-way handshake is never established, which makes the detection of the attack sender more difficult, and requires approaches able to identify probe attacks without connection stablishment. The UDP scan technique sends UDP packets to the destination port, and if it responds with a \emph{ICMP port unreachable} message, then it indicates that the scanned port is closed. On the other hand, if a message is not received, then the port is considered as open.

@@

\subsection{The DARPA Dataset}
\label{sec:DarpaDataset}

The DARPA 1998 dataset\footnote{https://www.ll.mit.edu/ideval/data/} includes 7 weeks of sniffed traffic saved into raw TCPDUMP packet data, from inside and outside origins, with labeled attacks. The attacks in this dataset
can be grouped into: denial-of-service (DoS); remote to local (R2L), which is characterized by unauthorized access from a
remote machine; user to root (U2R), which is characterized by unauthorized access to local
super-user privileges; and probe attack. Since the proposed approach focus on flood and probe attack, we analyze the attacks that present behaviors similar to flood or probe attack. We observe that the most cases of DoS focus on exploit system vulnerabilities instead of on flooding attack. One example is the occurrence of a neptune attack which sends 20 only SYN packets, what is a behavior that differs of the expected flood attack behavior. Therefore, we select the cases that simulates several network traffic or numerous connection requests, also known as flooding attack \cite{ahmed2016survey,osanaiye2016distributed}, and the cases that scan ports sending just a few packets. From the simulated probe attacks, we select the cases that rely on TCP or UDP connections.

The data modeling follows the method described by the Subsection \ref{sec:ModelingData}, with time frames of 20 minutes, packet counting aggregation by minute and considering the traffic to the following ports: 20, 21, 22, 23, 25, 79, 80, 88, 107, 109, 110, 113, 115, 143, 161, 389, 443.

@@

\begin{figure}[h!]
	\centering
     \includegraphics[height=11cm, width=9cm]{results/figures/mos_eigen_similarity.eps}
     \caption{Overview of The Framework for Detection and Identification of Network Attacks.}
     \label{fig:fig80}
\end{figure}

@@

According to flood and port scan attacks' behavior, flood attacks and port scan attacks can be characterized as covariance aware attack \citep{jin2004covariance} and correlation aware attack \citep{lakhina2005mining}, respectively. These characteristics are substantiated by the results obtained through the analysis based on sample covariation of zero mean variables and on covariance of zero mean and unitary standard deviation variables, described in Section \ref{sec:experimentalresults}, which shows that the main components of flood attacks are dominated by the variables with more variance and that the traffic associated with port scan attack does not generate many logs, however, it presents high covariance of zero mean and unitary standard deviation variables.

Therefore, to detect flood attacks, it is necessary to calculate the sample covariance matrix $\boldsymbol{\hat{R}}_{yy}^{(q)}$ of the zero mean samples given by

@@

Once the $\boldsymbol{\hat{R}}_{yy}^{(q)}$ and $\boldsymbol{\hat{R}}_{zz}^{(q)}$ have been obtained for flood and port scan attack detection, respectively, and since the next steps are the same for both sample covariance matrices, we refer to $\boldsymbol{\hat{R}}_{yy}$ and $\boldsymbol{\hat{R}}_{zz}$ as a matrix $\boldsymbol{C}$. Therefore, the following step of the algorithm is the eigenvalue decomposition (EVD), calculated according to (\ref{eq:eq06}), in order to obtain the vector of eigenvalues $\boldsymbol{e}^{(q)}$ associated with each matrix, according to (\ref{eq:eq060}).

\begin{equation}\label{eq:eq06}
\boldsymbol{C}^{(q)} = \boldsymbol{V}^{(q)}\boldsymbol{\Lambda}^{(q)}\boldsymbol{V}^{(q)^{\rm T}},
\end{equation}

\begin{equation}\label{eq:eq060}
\boldsymbol{e}^{(q)} = \rm diag(\boldsymbol{\Lambda}^{(q)}),
\end{equation}

@@

\begin{equation}\label{eq:eq11}
s_n = \frac{\abs{\boldsymbol{v}^{(q)} \cdot \boldsymbol{v}_{(n)}}}{\norm{\boldsymbol{v}^{(q)}}\norm{\boldsymbol{v}_{(n)}}},
\end{equation}
where $s_n$ denotes the absolute similarity degree of the $n$-th minute, $\boldsymbol{v}^{(q)}$ is the most significant eigenvectors of a selected set of minutes without network attack, and $\boldsymbol{v}_{(n)}$ is the most significant eigenvectors obtained after append the target $n$-th minute of traffic to be performed the flood and port scan attack identification.

@@

This section presents the performed experiments and the acquired results. First, in Section \ref{sec:AnalyzedScenario}, the scenario adopted in the experiments is summarized. Then, Section \ref{sec:largesteigenvaluesanalysis} shows the results of the largest eigenvalue analysis by time frames. In Section \ref{sec:MOSSchemesEvaluation} are described the results of the evaluated MOS schemes for attack detection. Section \ref{sec:EigenvalueAnalysis} presents the results of the eigenvalue analysis for identification of time frames under attack, Section \ref{sec:EigenSimilarityAnalysis} shows the results of similarity analysis for detailed flood and port scan identification, and Section \ref{sec:Complexity} discuss the computational complexity of the proposed framework.

@@

For the evaluation of MOS Schemes accuracy for flood and port scan detection, the framework defines that it is necessary to obtain the largest eigenvalue of each time frame, through eigen decomposition from a covariance of zero mean variables or covariance matrix of zero mean and unitary standard deviation variables, calculated from the evaluated traffic, as described in Section \ref{sec:prop_getv}. Through eigenvalue analysis of traffic with flood or port scan attacks, it is possible to visualize a significant difference between the largest eigenvalues and the remain eigenvalues, which can indicate a relationship between an outlier and time frames under attack.

@@

These results highlight that all $q$-th time frames where a network attack was simulated, present high significant variance between the largest eigenvalue and the remaining eigenvalues, obtained from sample covariance matrix, for flood detection, or from covariance matrix of zero mean and unitary standard deviation variables, for port scan detection. Therefore, we propose to apply the vector of the largest eigenvalues to MOS schemes in order to evaluate their accuracy for identification of time frames under attack, motivated by the fact that it is relevant to apply MOS schemes to automate the attack detection process, taking into account the characteristics of the evaluated eigenvalues.

@@

\begin{table}[h!]
  \centering
  \scriptsize
  \caption{MOS schemes applied to port scan and flood detection}
  \label{tab:tab4}
  \begin{tabular}{ c c c c c c c c }
	\toprule
	\multirow{2}{*}{\textbf{Type of analysis} $q$} &\multicolumn{6}{c}{\textbf{MOS schemes (estimated model order $\hat{d}$)}} &{\textbf{(d)}}\\ 
			\hhline{~------~}
		&\textbf{AIC} &\textbf{MDL} &\textbf{EDC} &\textbf{RADOI} &\textbf{EFT} &\textbf{SURE}\\
	\midrule
	Detection of synflood \\(presence of attack) &2 &1 &\textbf{1} &5 &\textbf{1} &4 &\textbf{1} \\
	Detection of synflood \\(absence of attack) &1 &1 &\textbf{0} &1 &\textbf{0} &3 &\textbf{0} \\
	\midrule
	Detection of fraggle \\(presence of attack) &1 &1 &\textbf{1} &5 &\textbf{1} &4 &\textbf{1} \\
	Detection of fraggle \\(absence of attack) &1 &1 &\textbf{0} &1 &\textbf{0} &3 &\textbf{0} \\
	\midrule
	Detection of port scan \\(presence of attack) &1 &1 &\textbf{1} &1 &\textbf{1} &9 &\textbf{1} \\
	Detection of port scan \\(absence of attack) &0 &0 &\textbf{0} &1 &\textbf{0} &1 &\textbf{0} \\
	\midrule
	Detection of synflood/fraggle \\(presence of attack) &2 &2 &\textbf{2} &5 &\textbf{2} &5 &\textbf{2} \\
	Detection of synflood/fraggle \\(absence of attack) &1 &1 &\textbf{0} &1 &\textbf{0} &3 &\textbf{0} \\
    \bottomrule
  \end{tabular}
\end{table}

@@

In Table \ref{tab:tab4}, two MOS schemes outperforms from the others, EDC and EFT. Efficient Detection Criterion (EDC) and Exponential Fitting Test (EFT) are the most effective schemes, correctly estimating the number of attacks in comparison to the expected values for effetctive attack detection, as defined by the column of real values in Table \ref{tab:tab4}. The AIC and MDL schemes are satisfactory only for port scan detection, however SURE and RADOI schemes did not show effective results for port scan or flood detection.

Although EDC and EFT presented the same accuracy on the evaluation, the EDC scheme requires less processing time than EFT, which is an important criteria to select EDC as the MOS scheme for flood and port scan detection on the remain experiments.

@@

These results highlight the capability of change detection based on similarity between legitimate and malicious traffic from flood or port scan attacks, endorsing the effectiveness and safety for adoption of threshold for attack detection through eigen similarity analysis.

@@

Given $\hat{N}$, which is the set of estimated $n$-th minutes under attack, it is possible to apply cosine similarity analysis to identify variation of the most significant eigenvectors, caused by the insertion of anomalous network traffic by a selected $m$-th port, during a $n$-th minute. Therefore, the incremental individualized and individual approaches of eigen similarity analysis were evaluated, for detection of ports under flood and port scan attacks, according to results presented in following tables. For this evaluation, the $v$ last most significant eigenvectors without attack was used as reference for similarity analysis against each target port $m$-th.

@@

According to the shown significant lack of similarity between legitimate and malicious traffic, it is possible to adopt safe thresholds for flood and port scan detection through eigen similarity analysis.

@@

This paper evaluated the effectiveness of MOS schemes for fraggle attack detection, extending our previous work \cite{tenorio2013greatest} and showing that the analysis of the largest eigenvalues by time frames can be applied to detect the number of port scanning, and flood attacks, but still requiring more information for detailed attack detection. Therefore, we proposed a novel approach for detailed network attack detection, based on eigen similarity analysis.

@@

According to the significant similarity difference between legitimate and malicious traffic, it is possible to adopt safe thresholds for flood and port scan detection through eigen similarity analysis.
